### 2023-09-28 chapters/background.tex
Work on discussion and conclusion, plotting

\subsection{Variational autoencoders} 
To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb,\zb)$ with a neural network with parameters $\thetab$. Specifically, they use a neural network to parameterize the conditional distribution $p_\thetab(\xb|\zb)$ and a simple distribution for the prior $p(\zb)$. 
Since this makes the marginal and posterior distributions intractable, both training and inference require approximate methods. 
While MCMC methods are computationally expensive for large neural networks, variational inference provides a more scalable alternative by estimating the posterior distribution $p(\zb|\xb)$ with a variational approximation.

@@

\paragraph{Applications of VAEs}
Despite the challenges facing the training of VAEs, they have been successfully applied to a number of tasks including image generation \cite{kingma_autoencoding_2014,rezende_stochastic_2014}, image inpainting \cite{pathak_context_2016}, image super-resolution \cite{sonderby_amortised_2017,chira_image_2022}, and speech synthesis \cite{hsu_unsupervised_2017,hsu_hierarchical_2019}. 
VAEs have also proven themselves useful for semi-supervised learning \cite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}. Recent architectural advances have made it possible to train VAEs with a large number of latent variables \cite{sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021} and have alleviated the posterior collapse problem \cite{maaloe_biva_2019}.

\paragraph{Self-supervised learning}
An alternative approach to representation learning is provided by self-supervised learning \cite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself. 
While it is fair to say that not all self-supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the field of natural language processing \cite{devlin_bert_2018,chen_simple_2020}, speech processing \cite{schneider_wav2vec_2019}, and computer vision \cite{chen_simple_2020}. 
For a deeper comparison of VAEs and self-supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}.

>>>>

\subsection{Variational autoencoders} \label{sec:variational-autoencoders}
To model high-dimensional data, \textcite{kingma_autoencoding_2014,rezende_stochastic_2014} propose to parameterize the generative model $p_\thetab(\xb,\zb)$ with a neural network with parameters $\thetab$. Specifically, they use a neural network to parameterize the conditional distribution $p_\thetab(\xb|\zb)$ and a simple distribution for the prior $p(\zb)$. 
Since this makes the marginal and posterior distributions intractable, both training and inference require approximate methods. 
While MCMC methods are computationally expensive for large neural networks, variational inference provides a more scalable alternative by estimating the posterior distribution $p(\zb|\xb)$ with a variational approximation.

@@

\paragraph{Applications of VAEs}
Despite the challenges facing the training of VAEs, they have been successfully applied to a number of tasks including image generation \cite{kingma_autoencoding_2014,rezende_stochastic_2014}, image inpainting \cite{pathak_context_2016}, image super-resolution \cite{sonderby_amortised_2017,chira_image_2022}, and speech synthesis \cite{hsu_unsupervised_2017,hsu_hierarchical_2019}. 
VAEs have also proven themselves useful for semi-supervised learning \cite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}. Recent architectural advances have made it possible to train VAEs with many latent variables \cite{sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021} and have alleviated the posterior collapse problem \cite{maaloe_biva_2019}.

\paragraph{Self-supervised learning}
An alternative approach to representation learning is provided by self-supervised learning \cite{mikolov_efficient_2013,devlin_bert_2018,chen_simple_2020,schneider_wav2vec_2019} which is a form of unsupervised learning where the training objective is derived from the data itself. 
While it is fair to say that not all self-supervised training objectives are as well-motivated as the ELBO is for VAEs, they have shown impressive results within the field of natural language processing \cite{devlin_bert_2018,chen_simple_2020}, speech processing \cite{schneider_wav2vec_2019}, and computer vision \cite{chen_simple_2020}. 
For a deeper comparison of VAEs and self-supervised methods for speech processing, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}. 
\Cref{chp:paper-review} provides a comprehensive introduction to and review of self-supervised learning in speech recognition.

### 2023-09-28 chapters/conclusion.tex
Work on background
Work on discussion and conclusion, plotting

\chapter[conclusions and outlook]{Conclusions and outlook}\label{chp:conclusion}

% ~3 pages

A conclusion with some abbreviations and citations, \gls{VAE} \cite{kingma_autoencoding_2014,rezende_stochastic_2014}.

\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}

>>>>

\chapter[conclusions and outlook]{Conclusions and outlook}\label{chp:conclusion}
% ~3 pages

In the introductory \cref{chp:introduction} we laid out how, through human history, progress-driving technological development has always introduced new challenges and risks of error whether via misuse, misunderstanding or inherent limitations. We emphasized that machine learning comes with these same types of risk may in some cases may amplify their impacts and argued that uncertainty estimation is a key component in developing safe and reliable machine learning systems. We provided the motivational cases of recognition of stroke cases in emergency calls \cref{subsec:motivation-stroke-recognition} and automated medical coding \cref{subsec:motivation-medical-coding} and gave examples of potential sources of uncertainty and of how correctly quantifying uncertainty can improve the usefulness of such tools in practice. Finally, we ended the introduction by connecting machine learning reliability \cref{sec:machine-learning-reliability} and model calibration \cref{subsec:model-calibration} and defining different types of uncertainty \cref{subsec:types-of-uncertainty}.

\Cref{chp:technical-background} provided technical background only covered briefly by the individual studies. We first introduced uncertainty as a concept in the context of information and probability theory \cref{sec:information-theory}. Then, we defined the task of out-of-distribution detection and reviewed existing work on the problem \cref{\cref{sec:out-of-distribution-detection}}. Finally, we provided technical background for variational autoencoders \cref{sec:variational-autoencoders}.

\dots

### 2023-09-28 chapters/discussion.tex
Work on discussion and conclusion, plotting

\chapter[discussion]{Discussion}\label{chp:discussion}

% ~5 pages

@@

\lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other). Include reference to \cite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
\lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimzing a mutual information term in the ELBO (derive this form).}
\lesstodo[inline]{Discuss some recent work e.g. \cref{morningstar_density_2021}}

@@

\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation? \cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}

\section{\Cref{chp:paper-retrospective} revisited: \dots}

\lesstodo[inline]{Discuss the calibration of the stroke model and report calibration curve.}
\lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html

\lesstodo[inline]{Maybe mention MultiQT paper \cite{havtorn_multiqt_2020}.}

>>>>

\chapter[discussion]{Discussion}\label{chp:discussion}
% ~5 pages

@@

\lesstodo[inline]{Discuss sensitivity to ``implicit" prior such as architecture (and probably optimization method and other). Include reference to \cite{huszar_is_2017} discussing the usefulness of using a maximum likelihood objective for representation learning in generative models.}
\lesstodo[inline]{Discuss whether VAEs are even suitable for representation learning due to them minimizing a mutual information term in the ELBO (derive this form).}
\lesstodo[inline]{Discuss some recent work e.g. \cref{morningstar_density_2021}}

@@

\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation? \cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}
% Since the main focus within self-supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self-supervised representations in terms of uncertainty estimation. 
% However, in the context of medical applications where data can be abundant but labels are sparse, unsupervised uncertainty estimation is a very interesting direction for future work.
\lesstodo[inline]{OOD data: Generalization or detection (https://arxiv.org/pdf/2110.11334.pdf)?}

\section{\Cref{chp:paper-retrospective} revisited: Uncertainty}
The work in \textcite{wenstrup_retrospective_2023} deals with predictive performance and analyses feature importance but does not explicitly consider uncertainty estimation.

\begin{figure}
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{python_plotting/calibration_curve_ensemble_and_all_models_uncalibrated.pdf}
        \caption{}
        \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{python_plotting/histogram_ensemble_and_single_model.pdf}
        \caption{}
        \label{fig_hierarchical:histogram_ensemble_and_single_model}
    \end{subfigure}
\end{figure}

\lesstodo[inline]{Discuss the calibration of the stroke model and report calibration curve.}
\lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html
\lesstodo[inline]{Maybe mention MultiQT paper \cite{havtorn_multiqt_2020}.}

### 2023-09-28 chapters/introduction.tex
Work on discussion and conclusion, plotting

\begin{figure}[t]
    \centering
    % \includegraphics[width=0.90\textwidth]{plimpton_322.png}
    \includegraphics[height=0.4\textheight]{plimpton_322.png}
    \caption{Technology is difficult. The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics. \cite[photo credit]{neugebauer_mathematical_1945}}
    \label{fig:plimpton_332}
\end{figure}

@@

\begin{figure}[t]
    \centering
    % \includegraphics[width=0.95\textwidth]{eniac_programmers.jpeg}
    \includegraphics[height=0.4\textheight]{eniac_programmers.jpeg}
    \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC. \cite[photo credit]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}} %Photo: \copyright Corbis/ Getty Images.}
    \label{fig:eniac_programmers}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=0.95\textwidth]{uber_nhsa_accident.png}
    \caption{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018. The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \cite[photo credit]{nationaltransportationsafetyboardnhsa_collision_2019}.}
    % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N. Mill Avenue and movement and speed of SUV at three points before impact. Pedestrian's path shows her position from initial detection ($5.6$ seconds before impact) until impact; SUV's position is shown at corresponding times beginning $4.2$ seconds before impact \cite{nationaltransportationsafetyboardnhsa_collision_2019}.}
    \label{fig:uber_nhsa_accident}
\end{figure}

@@

\subsection{Stroke recognition in emergency calls}
%
\paragraph{Motivation} In 2021, Corti entered a research collaboration with the Capital Region of Denmark and the Copenhagen Emergency Services to develop a system for recognizing stroke cases in calls to the 1-1-2 emergency line and the 1813 medical helpline. 
Stroke is one of the biggest causes of disability and death worldwide \cite{cite1,cite2,cite3} and effective treatment is highly time-sensitive \cite{cite4,cite5}. The most common gateway to specialized treatment and hospital admittance is through prehospital telehealth services like emergency medical call centers, nurse advice call lines, and out-of-hours health services \cite{cite6,cite7}, however, studies have found that approximately half of all patients with stroke do not receive the correct triage for their condition from call-takers \cite{cite10,cite11,cite12}. 
This is likely due to the complexity of stroke cases which exhibit a wide range of symptoms that can be difficult to recognize over the phone. Additionally, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls made to the Copenhagen 1-1-2 emergency line in 2021. This makes it difficult for call-takers to gain practical experience with stroke cases which compounds to the difficulty of recognizing them. Although several efforts have been made to improve recognition rates, there is still a need for better tools to support call-takers in recognizing stroke cases \cite{cite13,cite14,cite15}.

@@

\subsection{Medical coding of clinical notes}
%
\paragraph{Motivation} During the course of this project, Corti's portfolio has expanded to include a software system for medical coding of clinical notes. 
When a patient is admitted to a hospital, the medical staff will write a clinical note describing the patient's diagnosis and the procedures they underwent, including drug prescriptions. 
These notes are then used for billing and reimbursement purposes, as well as for research and quality assurance. 
The clinical notes are written, usually by a doctor, in natural language adhering to a certain structure. Later, a medical coder will assign a set of codes to the note based on its content. 
The process of medical coding is time-consuming and error-prone due to the vast number and high complexity of medical diagnoses and procedures. For instance, the widely used International Classification of Diseases (ICD) standard consist of 55,000 medical codes in version 10 and 85,000 in version 11 \cite{worldhealthorganisationwho_international_2023}. Additionally, a single medical note will usually contain several diagnoses and procedures which must all be inferred from the clinical note \cite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}. 
Furthermore, any single code can have several exclusion or inclusion criteria defined by official guidelines that determine when other codes must be coded along with it, and when they must not \cite{centersformedicaremedicaidservicesus_icd10cm_2023}. 
% Official guidelines describe how to code specific diagnoses and procedures and combinations thereof . 
% For instance, a code may have several exclusion or inclusion criteria which define which other codes must be coded along with it, or when they must not.

@@

\section{Machine learning reliability}
%
Several factors contribute to the reliability of a machine learning model. 
Besides model performance and accuracy, important factors include the interpretability of how the model functions, the explainability of its predictions, fairness in its treatment of different groups, and robustness to noise, outliers and adversarial attacks. 
% In medical applications, it has been argued that another factor
Since many modern machine learning models are deep neural networks with millions or billions of parameters, their size and complexity make them inherently difficult to interpret, and their predictions hard to explain. Due to high cost and practical infeasibility, the vast amounts of data needed to train such models are often not manually curated for quality, and so, may contain biases and errors which models are well-equipped to learn to mirror \cite{burkart_survey_2021}. 
Finally, a number of factors, including the ability of deep neural networks to overfit, or even memorize, the training data \cite{arpit_closer_2017, burg_memorization_2021}, can lead to models that are sensitive adverse noise conditions and outliers.

@@

In practice, miscalibrated models are not reliable in high risk applications. Although no model is perfect, the inability of a human, or another model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process. 
% A connection can be made back to the incident with the Uber self-driving car. 
The multitude of rapid misclassifications made by the Uber self-driving car in Tempe is likely an indication that the model responsible for object-classification was overconfident in its predictions and unable to represent or communicate uncertainty to other systems in the car, or the safety driver. Instead, the model's predictions were taken at face value and subsystems in the car acted on them without appropriate consideration to their reliability.

\subsection{Model calibration}
% 
The problem of miscalibration in machine learning models is well-known and has been studied for decades \cite{lewis_sequential_1995, platt_probabilistic_1999, garczarek_classification_2002, zadrozny_transforming_2002, bennett_using_2003}. One of the best known methods for calibrating a binary classifier is Platt scaling which fits a logistic regression model to the model outputs, assuming that the miscalibration can be corrected by a logarithmic function \cite{platt_probabilistic_1999}. Another method is isotonic regression which instead fits a non-parametric, monotonic function \cite{zadrozny_transforming_2002}. 
More recently, \textcite{guo_calibration_2017} proposed a single-parameter variant of Platt-scaling that fits only a temperature parameter on the logits of a neural network classifier. 
These methods are generally simple and effective, but not without limitations. 
To perform the calibration, most of the methods require a held-out validation set on which the model was not trained, Platt scaling and isotonic regression are not naively applicable to multi-class classification problems, and how to calibrate models for structured prediction tasks, such as speech recognition and machine translation, remains an open problem \cite{astudillo_uncertainty_2010, astudillo_integration_2013, jayashankar_detecting_2020}.

@@

\subsection{Understanding uncertainty}
% 
% OUTLINE: Introduce aleatoric and epistemic uncertainty. 
% - Discuss the distinction between them and how they relate to the problem of uncertainty estimation in machine learning.
% - 
% 
As hinted at in the previous section, there are different sources of uncertainty in machine learning models. 
At a fundamental level, we can distinguish between uncertainty that is present in the knowledge we have, and uncertainty that originates from the knowledge that we do not have.
These sources are sometimes called \emph{known unknowns} and \emph{unknown unknowns}, or referred to in terms of \textit{aleatoric uncertainty} and \textit{epistemic uncertainty} \cite{kendall_what_2017}. 
Aleatoric comes from the Latin word 'aleatorius' for 'dice player' or 'gambler' and refers to uncertainty present in the data itself due to randomness in the process that generated it. This kind of uncertainty is commonly seen as irreducible but, since it is represented in the collected data, it can usually be modelled directly. 
Epistemic comes from the Greek word 'epistḗmē' which means 'knowledge' and refers to uncertainty due to things one could in principle know, but does not in practice. Such lack of knowledge could for instance be due to a lack of data, or an improper model specification. 
Aleatoric uncertainty is sometimes referred to as \textit{stochastic uncertainty} and epistemic uncertainty as \textit{systematic uncertainty} \cite{kendall_what_2017}. 
Although the distinction between aleatoric and epistemic uncertainty is important, it is not always useful. For instance, a model may be uncertain about a particular prediction due to any of these two kinds of uncertainty. For that reason, the quantity of interest is usually the combined aleatoric and epistemic uncertainties, often referred to as \textit{predictive uncertainty}.

@@

Since aleatoric uncertainty is present in the data, it can usually be modelled directly. For instance, in the case of speech recognition, we can model the uncertainty in the transcription of a word by a distribution over the words in the vocabulary. But how do we model epistemic uncertainty? 
Since sources of epistemic uncertainty are by definition those not represented in the data, it is not generally possible to model them directly. This makes epistemic uncertainty more difficult to quantify than aleatoric uncertainty. 
Methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995} which represent it by an explicit distribution over learned model parameters. Ensembling methods \cite{gal_dropout_2016,lakshminarayanan_simple_2017} take a similar approach but use an implicit distribution. 
Other approaches include anomaly detection and the recent field of out-of-distribution detection which for example represent epistemic uncertainty by special output classes for out-of-distribution data, distributions over data representations, or distances between them (see \cref{sec:out-of-distribution-detection}).

@@

\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. \Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to estimation of epistemic uncertainty and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}. \Cref{part:self-supervised-speech-representation-learning} contains a review paper on self-supervised speech representation learning \cite{mohamed_selfsupervised_2022}. \Cref{part:medical-applications} contains papers on applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.

>>>>

\begin{figure}[t]
    \centering
    % \includegraphics[width=0.90\textwidth]{plimpton_322.png}
    \includegraphics[height=0.4\textheight]{plimpton_322.png}
    \caption{Technology is difficult. The scribe of the Plimpton 322 clay tablet ($\sim$1800 BCE) made several errors when listing Pythagorean triplets, presumably struggling to grapple with Babylonian mathematics. \cite[photo credit][]{neugebauer_mathematical_1945}}
    \label{fig:plimpton_332}
\end{figure}

@@

\begin{figure}[t]
    \centering
    % \includegraphics[width=0.95\textwidth]{eniac_programmers.jpeg}
    \includegraphics[height=0.4\textheight]{eniac_programmers.jpeg}
    \caption{Ruth Lichterman (left) and Marlyn Wescoff (middle) were two of the several female programmers of the ENIAC. \cite[photo credit][]{usarmyresearchlaboratoryarltechnicallibrary_female_1940}} %Photo: \copyright Corbis/ Getty Images.}
    \label{fig:eniac_programmers}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=0.95\textwidth]{uber_nhsa_accident.png}
    \caption{A pedestrian was killed by an Uber self-driving car in Tempe, Arizona in 2018. The car's sensors detected the pedestrian $5.6$ seconds before the crash, but the self-driving system failed to recognize its own uncertainty in multiple object misclassifications and so did not correctly predict her path or reduce the SUV's speed \cite[photo credit][]{nationaltransportationsafetyboardnhsa_collision_2019}.}
    % \caption{Aerial view of crash location showing path of pedestrian as she attempted to cross N. Mill Avenue and movement and speed of SUV at three points before impact. Pedestrian's path shows her position from initial detection ($5.6$ seconds before impact) until impact; SUV's position is shown at corresponding times beginning $4.2$ seconds before impact \cite{nationaltransportationsafetyboardnhsa_collision_2019}.}
    \label{fig:uber_nhsa_accident}
\end{figure}

@@

\subsection{Stroke recognition in emergency calls} \label{subsec:motivation-stroke-recognition}
%
\paragraph{Motivation} In 2021, Corti entered a research collaboration with the Capital Region of Denmark and the Copenhagen Emergency Services to develop a system for recognizing stroke cases in calls to the 1-1-2 emergency line and the 1813 medical helpline. 
Stroke is one of the biggest causes of disability and death worldwide \cite{cite1,cite2,cite3} and effective treatment is highly time-sensitive \cite{cite4,cite5}. The most common gateway to specialized treatment and hospital admittance is through prehospital telehealth services like emergency medical call centers, nurse advice call lines, and out-of-hours health services \cite{cite6,cite7}, however, studies have found that approximately half of all patients with stroke do not receive the correct triage for their condition from call-takers \cite{cite10,cite11,cite12}. 
This is likely due to the complexity of stroke cases which exhibit a wide range of symptoms that can be difficult to recognize over the phone. Additionally, stroke cases are relatively rare, occurring in only $0.25\%$ of all calls made to the Copenhagen 1-1-2 emergency line in 2021. This makes it difficult for call-takers to gain practical experience with stroke cases which compounds to the difficulty of recognizing them. Although several efforts have been made to improve recognition rates, there is still a need for better tools to support call-takers in recognizing stroke cases \cite{cite13,cite14,cite15}.

@@

\subsection{Medical coding of clinical notes} \label{subsec:motivation-medical-coding}
%
\paragraph{Motivation} During the course of this project, Corti's portfolio has expanded to include a software system for medical coding of clinical notes. 
When a patient is admitted to a hospital, the medical staff will write a clinical note describing the patient's diagnosis and the procedures they underwent, including drug prescriptions. 
These notes are then used for billing and reimbursement purposes, as well as for research and quality assurance. 
The clinical notes are written, usually by a doctor, in natural language adhering to a certain structure. Later, a medical coder will assign a set of codes to the note based on its content. 
The process of medical coding is time-consuming and error-prone due to the vast number and high complexity of medical diagnoses and procedures. For instance, the widely used International Classification of Diseases (ICD) standard consist of 55,000 medical codes in version 10 and 85,000 in version 11 \cite{worldhealthorganisationwho_international_2023}. Additionally, a single medical note will usually contain several diagnoses and procedures which must all be inferred from the clinical note \cite{johnsonMIMICIIIFreelyAccessible2016,johnsonMIMICIVFreelyAccessible2023}. 
Furthermore, any single code can have several exclusion or inclusion criteria defined by official guidelines that determine when other codes must be coded along with it, and when they must not \cite{centersformedicaremedicaidservicesus_icd10cm_2023}. 
% Official guidelines describe how to code specific diagnoses and procedures and combinations thereof . 
% For instance, a code may have several exclusion or inclusion criteria which define which other codes must be coded along with it, or when they must not.

@@

\section{Machine learning reliability} \label{sec:machine-learning-reliability}
%
Several factors contribute to the reliability of a machine learning model. 
Besides model performance and accuracy, important factors include the interpretability of how the model functions, the explainability of its predictions, fairness in its treatment of different groups, and robustness to noise, outliers and adversarial attacks. 
% In medical applications, it has been argued that another factor
Since many modern machine learning models are deep neural networks with millions or billions of parameters, their size and complexity make them inherently difficult to interpret, and their predictions hard to explain. Due to high cost and practical infeasibility, the vast amounts of data needed to train such models are often not manually curated for quality, and so, may contain biases and errors which models are well-equipped to learn to mirror \cite{burkart_survey_2021}. 
Finally, a number of factors, including the ability of deep neural networks to overfit, or even memorize, the training data \cite{arpit_closer_2017, burg_memorization_2021}, can lead to models that are sensitive adverse noise conditions and outliers.

@@

In practice, miscalibrated models are not reliable, especially in high risk applications. Although no model is perfect, the inability of a human, or another model, to assess the certainty with which a prediction was made, makes it impossible to know how to weigh it in a decision-making process. 
% A connection can be made back to the incident with the Uber self-driving car. 
The multitude of rapid misclassifications made by the Uber self-driving car in Tempe is likely an indication that the model responsible for object-classification was overconfident in its predictions and unable to represent or communicate uncertainty to other systems in the car, or the safety driver. Instead, the model's predictions were taken at face value and subsystems in the car acted on them without appropriate consideration to their reliability.

\subsection{Model calibration} \label{subsec:model-calibration}
% 
The problem of miscalibration in machine learning models is well-known and has been studied for decades \cite{lewis_sequential_1995, platt_probabilistic_1999, garczarek_classification_2002, zadrozny_transforming_2002, bennett_using_2003}. One of the best known methods for calibrating a binary classifier is Platt scaling which fits a logistic regression model to the model outputs, assuming that the miscalibration can be corrected by a logarithmic function \cite{platt_probabilistic_1999}. Another method is isotonic regression which instead fits a non-parametric, monotonic function \cite{zadrozny_transforming_2002}. 
More recently, \textcite{guo_calibration_2017} proposed a single-parameter variant of Platt-scaling that fits only a temperature parameter on the logits of a neural network classifier. 
These methods are generally simple and effective, but not without limitations. 
To perform the calibration, most of the methods require a held-out validation set on which the model was not trained, Platt scaling and isotonic regression are not naively applicable to multi-class classification problems, and how to calibrate models for structured prediction tasks, such as speech recognition and machine translation, remains an open problem \cite{astudillo_uncertainty_2010, astudillo_integration_2013, jayashankar_detecting_2020}.

@@

\subsection{Understanding uncertainty} \label{subsec:understanding-uncertainty}
% 
% OUTLINE: Introduce aleatoric and epistemic uncertainty. 
% - Discuss the distinction between them and how they relate to the problem of uncertainty estimation in machine learning.
% - 
% 
As hinted at in the previous section, there are different sources of uncertainty in machine learning models. 
At a fundamental level, we can distinguish between uncertainty that is present in the knowledge we have, and uncertainty that originates from the knowledge that we do not have.
These sources are sometimes called \emph{known unknowns} and \emph{unknown unknowns}, or referred to in terms of \textit{aleatoric uncertainty} and \textit{epistemic uncertainty} \cite{kendall_what_2017}. 
Aleatoric comes from the Latin word 'aleatorius' for 'dice player' or 'gambler' and refers to uncertainty present in the data itself due to randomness in the process that generated it. This kind of uncertainty is commonly seen as irreducible but, since it is represented in the collected data, it can usually be modelled directly. 
Epistemic comes from the Greek word 'epistḗmē' which means 'knowledge' and refers to uncertainty due to things one could in principle know, but does not in practice. Such lack of knowledge could for instance be due to a lack of data, or an improper model specification. 
Aleatoric uncertainty is sometimes referred to as \textit{stochastic uncertainty} and epistemic uncertainty as \textit{systematic uncertainty} \cite{kendall_what_2017}. 
Although the distinction between aleatoric and epistemic uncertainty is important, it is not always useful. For instance, a model may be uncertain about a particular prediction due to any of these two kinds of uncertainty. For that reason, the quantity of interest is usually the combined aleatoric and epistemic uncertainties, often referred to as \textit{predictive uncertainty}.

@@

Since aleatoric uncertainty is present in the data, it can usually be modelled directly. For instance, in the case of speech recognition, we can model the uncertainty in the transcription of a word by a distribution over the words in the vocabulary. But how do we model epistemic uncertainty? 
Since sources of epistemic uncertainty are by definition those not represented in the data, it is not generally possible to model them directly. This makes epistemic uncertainty more difficult to quantify than aleatoric uncertainty. 
Methods for estimating epistemic uncertainty include Bayesian neural networks \cite{mackay_practical_1992, neal_bayesian_1995} which represent it by an explicit distribution over learned model parameters. Ensemble methods \cite{gal_dropout_2016,lakshminarayanan_simple_2017} take a similar approach but use an implicit distribution. 
Other approaches include anomaly detection and the recent field of out-of-distribution detection which, for example, represent epistemic uncertainty by special output classes for out-of-distribution data, distributions over data representations, or distances between them (see \cref{sec:out-of-distribution-detection}).

@@

\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. \Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \cite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}. \Cref{part:self-supervised-speech-representation-learning} contains a review paper on self-supervised speech representation learning \cite{mohamed_selfsupervised_2022}. \Cref{part:medical-applications} contains papers on applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \cite{wenstrup_retrospective_2023} and medical coding of clinical notes \cite{edin_automated_2023}.

### 2023-09-28 chapters/main-contributions.tex
Work on discussion and conclusion, plotting

\contribitem{part:background}
\Cref{chp:introduction} provides a general introduction to the thesis and gives motivating examples for speech recognition and assistance in medical diagnostics. \Cref{chp:technical-background} provides additional technical background not included in the individual studies by introducing uncertainty as a concept in the context of information and probability theory, introducing the task of out-of-distribution detection and reviewing existing work on the problem, and providing an introduction to variational autoencoders.

\contribitem{part:unsupervised-uncertainty-estimation}
This part of the thesis is concerned with unsupervised uncertainty estimation and consists of three papers. 
The first two papers are concerned with out-of-distribution detection using generative models. The first paper proposes a new method for out-of-distribution detection using hierarchical variational autoencoders. 
The second paper builds on the first and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable generative model with explicit likelihood. 
The third paper proposes a new hierarchical variational autoencoder for speech and benchmarks it against other variational autoencoders and autoregressive models.

>>>>

\contribitem{part:background}
\Cref{chp:introduction} provides a general introduction to the thesis and gives motivating examples for speech recognition and assistance in medical encounters including stroke recognition on medical helplines and automation of medical coding. \Cref{chp:technical-background} provides additional technical background not included in the individual studies by introducing uncertainty as a concept in the context of information and probability theory, introducing the task of out-of-distribution detection and reviewing existing work on the problem, and providing an introduction to variational autoencoders.

\contribitem{part:unsupervised-uncertainty-estimation}
This part of the thesis is concerned with unsupervised uncertainty estimation and consists of three papers. 
The first two papers are approach unsupervised uncertainty estimation through the task of out-of-distribution detection using generative models. The first paper proposes a new method for out-of-distribution detection using hierarchical variational autoencoders. 
The second paper builds on the first and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable generative model with explicit likelihood. 
The third paper proposes a new hierarchical variational autoencoder for speech and benchmarks it against other variational autoencoders and autoregressive models.
